---
title: "Basis expansion"
author: "Tedoiler"
date: "02/12/2021"
output: html_document
---

https://bobby.gramacy.com/surrogates/splines.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basis expansions

Suppose that we have the classical linear model $Y = f(x) + \varepsilon$

We aim to transform the data matrix $X$ to a higher dimension.
We create $h_m(x): \mathbb{R}^p \to \mathbb{R}$, the $m$-th transformation of $x$ for $m=1,\dots, M$
Thus we can substitute $f(x)$ with an function of the form.
$$
f(x) = \sum_{m=1}^M \beta_m h_m(x)
$$
This is a **linear basis expansion** of $x$

Consider that we have some data that are being generated from some underlying sinusoidal function.

```{r}
x <- seq(0,10,length=50)
y_true <- sin(pi*x/5) + 0.2*cos(4*pi*x/5)
y <- y_true + rnorm(length(y_true), sd=0.1)
plot(x,y)
lines(x, y_true)
```

##### Constant Basis
```{r}
knot_1 <- x[round(length(x)/3)]
knot_2 <- x[round(2*length(x)/3)]
xi <- c(knot_1,knot_2)
hc_1 <- function(x) {as.numeric(x < xi[1])}
hc_2 <- function(x) {as.numeric(x >= xi[1] & x < xi[2])}
hc_3 <- function(x) {as.numeric(x >= xi[2])}

H_c <- data.frame(cbind(hc_1(x), hc_2(x), hc_3(x)))
names(H_c) <- paste("hc", 1:3, sep="")
```
We now can fit
```{r}
pc_fit <- lm(y~.-1, data=H_c)
```

and create a predictive grid and the true output for comparison.
```{r}
xx <- seq (-1,11, length=199)
HH_constant <- data.frame(cbind(hc_1(xx), hc_2(xx), hc_3(xx)))
names(HH_constant) <- paste('hc', 1:3, sep="")
yytrue <- (sin(pi*xx/5) + 0.2*cos(4*pi*xx/5))
```

```{r}
pc <- predict(pc_fit, newdata=HH_constant, interval='prediction')
plot(x,y)
lines(xx, yytrue, col="blue")
lines(xx, pc[,1])
lines(xx, pc[,2], col=2, lty=2)
lines(xx, pc[,3], col=2, lty=2)
abline(v=xi, col="gray", lty=2)
```

##### Linear Basis
```{r}
H_linear <- cbind(H_c, hc_1(x)*x, hc_2(x)*x, hc_3(x)*x)
names(H_linear)[4:6] <- paste("hl", 1:3, sep="")
pl_fit <- lm(y~.-1, data=H_linear)
```

Create a predictive grid and the true output for comparison.
```{r}
HH_linear <- cbind(HH_constant, hc_1(xx)*xx, hc_2(xx)*xx, hc_3(xx)*xx)
names(HH_linear)[4:6] <- paste("hl", 1:3, sep="")
pl <- predict(pl_fit, newdata=HH_linear, interval='prediction')
```

```{r}
plot(x,y)
lines(xx, yytrue, col="blue")
lines(xx, pl[,1])
lines(xx, pl[,2], col=2, lty=2)
lines(xx, pl[,3], col=2, lty=2)
abline(v=xi, col="gray", lty=2)
```
This seams like a much better fit for the function. However, it still has a major issue. 
By examining the area around the knots, we see that the predicting function behaves weirdly.
Therefore, we need to enforce it to be **continuous**. 

#### Enforcing Continuity

We first create a ReLU function
```{r}
relu <- function(x){
  x[x<0] <- 0
  return(x)
}
```

And now we create a design from that basis
```{r}
H_lc <- data.frame(rep(1, length(x)), x, relu(x-xi[1]), relu(x-xi[2]))
names(H_lc) <- paste("hlc", 1:4, sep="")
H_lc
```
Fit the transformed data
```{r}
plc_fit <- lm(y~.-1, data=H_lc)
```

And make predictions
```{r}
HH_lc <- rep(1,length(xx))
HH_lc <- cbind(HH_lc, xx)
HH_lc <- cbind(HH_lc, relu(xx-xi[1]))
HH_lc <- data.frame(cbind(HH_lc, relu(xx-xi[2])))
names(HH_lc) <- paste("hlc", 1:4, sep="")
plc <- predict(plc_fit, newdata=HH_lc, interval="prediction")
```

```{r}
plot(x,y)
lines(xx, yytrue, col="blue")
lines(xx, plc[,1])
lines(xx, plc[,2], col=2, lty=2)
lines(xx, plc[,3], col=2, lty=2)
abline(v=xi, col="gray", lty=2)
```

##### Cubic Splines
```{r}
H_cs <- data.frame(cbind(rep(1,length(x)), x, x^2, x^3, relu(x-xi[1])^3, relu(x-xi[2])^3))
names(H_cs) <- paste("x", 1:6, sep="")
fit_cs <- lm(y~., data=H_cs)
```

Predict
```{r}
HH_cs <- data.frame(cbind(rep(1,length(xx)), xx, xx^2, xx^3, relu(xx-xi[1])^3, relu(xx-xi[2])^3))
names(HH_cs) <- paste("x", 1:6, sep="")
pcs <- suppressWarnings(predict(fit_cs, newdata=HH_cs, interval="prediction"))
```

Plot the results
```{r}
plot(x,y)
lines(xx, yytrue, col="blue")
lines(xx, pcs[,1])
lines(xx, pcs[,2], col=2, lty=2)
lines(xx, pcs[,3], col=2, lty=2)
abline(v=xi, col="gray", lty=2)
```
Bootstrap the results
```{r}
B <- 1000
YYpred.boot <- matrix(NA, nrow=B, ncol=length(xx))

for (b in 1:B){
  indices <- sample(1:length(x), length(x), replace=TRUE)
  xi <- x[indices]
  fit <- smooth.spline(xi, y[indices], df=11)
  YYpred.boot[b,] <- as.matrix(predict(fit, data.frame(xi=xx))$y)

}
```

```{r}
m <- apply(YYpred.boot, 2, mean)
q1 <- apply(YYpred.boot, 2, quantile, p=0.05)
q2 <- apply(YYpred.boot, 2, quantile, p=0.95)
```

```{r}
matplot(xx, t(YYpred.boot), type="l", lty=1, col="gray", xlab="y", ylab="y")
points(x, y)
lines(xx, m, lwd=2)
lines(xx, q1, col=2, lty=2)
lines(xx, q2, col=2, lty=2)
```

